Approximation as Differentiation

### 2025-03-31 notes/sec/biproduct.tex
secrefs.
Or coproducts.

\subsection{Biproducts}

@@

\begin{center}
\begin{tikzcd}
   A \arrow[r, "\inj_A", shift left] &
   A \biprod B \arrow[l, "\proj_A", shift left] \arrow[r, "\proj_B"', shift right] &
   B \arrow[l, "\inj_B"', shift right]
\end{tikzcd}
\end{center}

@@

\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \proj_A \comp \inj_A &= \id_A \\
   \proj_B \comp \inj_A &= \zero_{A,B}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \proj_B \comp \inj_B &= \id_B \\
   \proj_A \comp \inj_B &= \zero_{B,A}
\end{salign*}
\end{center}
\end{minipage}

@@

\begin{salign*}
\inj_A \comp \proj_A + \inj_B \comp \proj_B &= \id_{A \proj B}
\end{salign*}
\end{definition}

If $C$ has biproducts then it has products given by $(A \biprod B, \proj_A, \proj_B)$ and coproducts given by $(A
\biprod B, \inj_A, \inj_B)$.

\begin{proposition}
\item
\begin{itemize}
\item If $C$ has products (or coproducts) then they are biproducts.
\item If $C$ has a terminal (or initial) object then it is a zero object.
\end{itemize}
\end{proposition}

\noindent We say that $C$ is \emph{semi-additive} if it has finite biproducts.

>>>>

\subsection{Biproducts}
\label{sec:biproduct}

@@

\begin{center}
\begin{tikzcd}
   A \arrow[r, "\biinj_A", shift left] &
   A \biprod B \arrow[l, "\biproj_A", shift left] \arrow[r, "\biproj_B"', shift right] &
   B \arrow[l, "\biinj_B"', shift right]
\end{tikzcd}
\end{center}

@@

\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_A \comp \biinj_A &= \id_A \\
   \biproj_B \comp \biinj_A &= \zero_{A,B}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_B \comp \biinj_B &= \id_B \\
   \biproj_A \comp \biinj_B &= \zero_{B,A}
\end{salign*}
\end{center}
\end{minipage}

@@

\begin{salign*}
\biinj_A \comp \biproj_A + \biinj_B \comp \biproj_B &= \id_{A \biprod B}
\end{salign*}
\end{definition}

\begin{proposition}
\item
\begin{itemize}
\item If $C$ has biproducts then it has products given by $(A \biprod B, \biproj_A, \biproj_B)$ and coproducts
given by $(A \biprod B, \biinj_A, \biinj_B)$.
\item If $C$ has products (or coproducts) then they are biproducts.
\item If $C$ has a terminal (or initial) object then it is a zero object.
\end{itemize}
\end{proposition}

\noindent We say that $C$ is \emph{semi-additive} if it has finite biproducts. Having products or coproducts
is sufficient for semi-additivity:

\begin{proposition}
Suppose $C$ is $\CMon$-enriched.
\begin{itemize}
\item If $C$ has products given by $(A \times B, \proj_1, \proj_2)$ then:
\begin{enumerate}
\item $\prodM{0_{A,B}}{0_{A,C}} = 0_{A,B \times C}$
\item $\prodM{f_1 + g_1}{f_2 + g_2} = \prodM{f_1 + f_2}{g_1 + g_2}$
\item $(A \times B, \biinj_A, \biinj_B, \proj_1, \proj_2)$ is a
biproduct with $\biinj_A: A \to A \times B = \prodM{\id_A}{0_{A,B}}$ and $\biinj_B: B \to A \times B =
\prodM{0_{A,B}}{\id_B}$.
\end{enumerate}
\item If $C$ has coproducts given by $(A + B, \inj_1, \inj_2)$ then
\begin{enumerate}
\item $(A + B, \inj_1, \inj_2, \biproj_A, \biproj_B)$ is a
biproduct with $\biproj_A: A + B \to A = \coprodM{\id_A}{0_{B,A}}$ and $\biproj_B: A + B \to B =
\coprodM{0_{A,B}}{\id_B}$.
\end{enumerate}
\end{itemize}
\end{proposition}

\begin{proposition}
\label{prop:biproduct:prod-coprod}
Suppose $C$ a category with biproducts. Then the following commutes for any morphisms $f_1, g_1, f_2,
g_2$ in $C$:

\begin{center}
\begin{tikzcd}[row sep=3em, column sep=3em]
   & A \arrow[d, "\biinj_A", shift left] \arrow[dr, "f_2"] \\
   A' \arrow[r, "\prodM{f_1}{g_1}"] \arrow[dr, "g_1"'] \arrow[ur, "f_1"] & A \biprod B \arrow[u, "\biproj_A", shift left] \arrow[d, "\biproj_B"', shift right] \arrow[r, pos=0.35, "\coprodM{f_2}{g_2}"] & B' \\
   & B \arrow[u, "\biinj_B"', shift right] \arrow[ur, "g_2"']
\end{tikzcd}
\end{center}
\end{proposition}

### 2025-03-31 notes/sec/grothendieck.tex
Placeholder for FdVect.
Or coproducts.

\begin{proposition}
Suppose $C$ locally small. If $C$ has binary biproducts and set-indexed products, then $\Fam(C)$ is Cartesian
closed. \todo{Establish first that if $C$ has binary coproducts and set-indexed products,
then $\Fam(C)$ is symmetric monoidal, with the coproduct as monoidal product.}
\end{proposition}

\begin{proof}
Suppose $C$ is locally small, with binary biproducts given by $(\biprod,\inj,\proj)$ and set-indexed products
given by $(\prod,\eval_{\prod},\lambda_{\prod})$. For any objects $X, Y, Z$ in $\Fam(C)$ define:

\begin{enumerate}
\item $[X, Y]$ to be the object $(X',Y')$ in $\Fam(C)$ with $X' = \Fam(C)(X,Y)$ and $Y'_f =
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}}$ for every $f: X \to Y$.
\item The family of isomorphisms $\lambda_{X,Y,Z}$ (natural in $X$) sending any $(f, g): X \times Y \to Z$ to
$(f', g'): X \to [Y, Z]$ with:
\begin{itemize}
\item $f'(x) = (f \comp (x, -), \reindex{g}{(x,-)} \comp \inj_{\faml{Y}}):
Y \to Z$ where $\inj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g}{(x,-)} \comp \inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f})$ where $\inj_{\faml{X}_x}: \const_{\idx{Y}}(\faml{X}_x)
\to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\end{itemize}
\item The family of morphisms $\eval_{X,Y}: [X, Y] \times X \to Y$ in $\Fam(C)$ with
\begin{itemize}
\item $\idx{\eval_{X,Y}}(f, x) = \idx{f}(x)$
\item $\faml{\eval_{X,Y}}_{f, x} = \coprodM{{\eval_{\prod}}_x}{\faml{f}_x}:
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}} \biprod \faml{X}_x \to \faml{Y}_{\idx{f}(x)}$
\end{itemize}
\end{enumerate}

@@

To show $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f,g)$, suppose $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f',
g')$. Then:
\begin{itemize}
\item $f'(x) = (f^\dagger \comp (x, -), \reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{Y}}): Y \to Z$ where
$\inj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f^\dagger})$ where $\inj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$.
\end{itemize}
\noindent To show $f' = f : \idx{X} \to \Fam(C)(Y,Z)$, suppose $x \in \idx{X}$ and show $f'(x) = f(x): Y \to
Z$. Suppose $y \in \idx{Y}$.
\begin{enumerate}
\item $\idx{f'(x)}(y) = (f^\dagger \comp (x, -))(y) = f^\dagger(x,y) = \idx{f(x)}(y)$.
\item $\faml{f'(x)}_y = (\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{Y}})_y = \reindex{g^\dagger}{(x,-)}_y \comp
{\inj_{\faml{Y}}}_y = g^\dagger_{x,y} \comp {\inj_{\faml{Y}}}_y = \coprodM{{\eval_{\prod}}_y \comp
g_x}{\faml{f(x)}_y} \comp {\inj_{\faml{Y}}}_y = \faml{f(x)}_y$.
\end{enumerate}

To show $g' = g$, suppose $x \in \idx{X}$.
\begin{align*}
\small
g'_x
&=
\lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \inj_{\faml{X}_x}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{g^\dagger_{(x,y)} \comp (\inj_{\faml{X}_x})_y\}_{y \in \idx{Y}}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{\coprodM{{\eval_{\prod}}_y \comp g_x}{\faml{f(x)}_y} \comp (\inj_{\faml{X}_x})_y\}_{y \in \idx{Y}}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{{\eval_{\prod}}_y \comp g_x\}_{y \in \idx{Y}}) \\
&=
g_x
\end{align*}
\qed

@@

\begin{center}
\begin{tikzcd}
   \prod_{\idx{Y}}\reindex{\faml{Z}}{\idx{f} \comp (x,-)} \arrow[dr, "\eval_{\prod_{y}}"] \\
   \faml{X}_x \arrow[u, "\lambda_{\prod_{\idx{Y}}}(\reindex{\faml{f}}{{(x,-)}}\,\comp\,
\inj_{\faml{X}_x})"] \arrow[r, "\faml{f}_{x,y}\,\comp\,\inj_{\faml{X}_x}"'] & \faml{Z}_{\idx{f}(x,y)}
\end{tikzcd}
\hspace{1cm}
\begin{tikzcd}[column sep=5em]
   \faml{Y}_y \arrow[dr,
   "\faml{f}_{(x,y)}\,\comp\,\inj_{\faml{Y}_y}"] \\
   \faml{Y}_y \arrow[u, -, double, "\id_{\faml{Y}_y}"] \arrow[r, "\faml{f}_{x,y}\,\comp\,\inj_{\faml{Y}_y}"'] & \faml{Z}_{\idx{f}(x,y)}
\end{tikzcd}
\end{center}

The diagram on the left below then commutes by Proposition~\ref{prop:biproduct:prod-coprod}, with
$\coprodM{\faml{f}_{x,y}\,\comp\,\inj_{\faml{X}_x}}{\faml{f}_{x,y}\,\comp\,\inj_{\faml{Y}_y}} =
\faml{f}_{(x,y)}$ by the uniqueness of the coproduct of morphisms. Then the diagram on the right in $\Fam(C)$
commutes by the definitions of $\lambda_{X,Y,Z}, \eval_{Y,Z}$ and $[Y,Z]$.

\begin{center}
\begin{tikzcd}[column sep=5.5em]
   \prod_{\idx{Y}}\reindex{\faml{Z}}{\idx{f} \comp (x,-)} \biprod \faml{Y}_y \arrow[dr,
   "\coprodM{\eval_{\prod_{y}}}{\faml{f}_{(x,y)}\,\comp\,\inj_{\faml{Y}_y}}"] \\
   \faml{X}_x \biprod \faml{Y}_y \arrow[u, "(\lambda_{\prod_{\idx{Y}}}(\reindex{\faml{f}}{{(x,-)}}\,\comp\,
\inj_{\faml{X}_x}){,}\,\id_{\faml{Y}_y})"] \arrow[r,
"\coprodM{\faml{f}_{x,y}\,\comp\,\inj_{\faml{X}_x}}{\faml{f}_{x,y}\,\comp\,\inj_{\faml{Y}_y}}"'] &
\faml{Z}_{\idx{f}(x,y)}
\end{tikzcd}
\hspace{1cm}
\begin{tikzcd}
   {{[Y, Z]}} \tensor Y \arrow[dr, "\eval_{Y,Z}"] \\
   X \tensor Y \arrow[u, "\lambda_{X,Y,Z}(f) \tensor \id_Y"] \arrow[r, "f"'] & Z
\end{tikzcd}
\end{center}
\end{enumerate}
\end{proof}

>>>>

\begin{proposition}
Suppose $C$ locally small. If $C$ has binary biproducts and set-indexed products, then $\Fam(C)$ is Cartesian
closed.
\end{proposition}

\begin{proof}
Suppose $C$ is locally small, with binary biproducts given by $(\biprod,\biinj,\biproj)$ and set-indexed products
given by $(\prod,\eval_{\prod},\lambda_{\prod})$. For any objects $X, Y, Z$ in $\Fam(C)$ define:

\begin{enumerate}
\item $[X, Y]$ to be the object $(X',Y')$ in $\Fam(C)$ with $X' = \Fam(C)(X,Y)$ and $Y'_f =
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}}$ for every $f: X \to Y$.
\item The family of isomorphisms $\lambda_{X,Y,Z}$ (natural in $X$) sending any $(f, g): X \times Y \to Z$ to
$(f', g'): X \to [Y, Z]$ with:
\begin{itemize}
\item $f'(x) = (f \comp (x, -), \reindex{g}{(x,-)} \comp \biinj_{\faml{Y}}):
Y \to Z$ where $\biinj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g}{(x,-)} \comp \biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f})$ where $\biinj_{\faml{X}_x}: \const_{\idx{Y}}(\faml{X}_x)
\to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\end{itemize}
\item The family of morphisms $\eval_{X,Y}: [X, Y] \times X \to Y$ in $\Fam(C)$ with
\begin{itemize}
\item $\idx{\eval_{X,Y}}(f, x) = \idx{f}(x)$
\item $\faml{\eval_{X,Y}}_{f, x} = \coprodM{{\eval_{\prod}}_x}{\faml{f}_x}:
\prod_{\idx{X}}\reindex{\faml{Y}}{\idx{f}} \biprod \faml{X}_x \to \faml{Y}_{\idx{f}(x)}$
\end{itemize}
\end{enumerate}

@@

To show $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f,g)$, suppose $\lambda_{X,Y,Z}(f^\dagger, g^\dagger) = (f',
g')$. Then:
\begin{itemize}
\item $f'(x) = (f^\dagger \comp (x, -), \reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{Y}}): Y \to Z$ where
$\biinj_{\faml{Y}}: \faml{Y} \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$
\item $g'_x = \lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \reindex{\faml{Z}}{f^\dagger})$ where $\biinj_{\faml{X}_x}:
\const_{\idx{Y}}(\faml{X}_x) \to \const_{\idx{Y}}(\faml{X}_x) \biprod \faml{Y}$.
\end{itemize}
\noindent To show $f' = f : \idx{X} \to \Fam(C)(Y,Z)$, suppose $x \in \idx{X}$ and show $f'(x) = f(x): Y \to
Z$. Suppose $y \in \idx{Y}$.
\begin{enumerate}
\item $\idx{f'(x)}(y) = (f^\dagger \comp (x, -))(y) = f^\dagger(x,y) = \idx{f(x)}(y)$.
\item $\faml{f'(x)}_y = (\reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{Y}})_y = \reindex{g^\dagger}{(x,-)}_y \comp
{\biinj_{\faml{Y}}}_y = g^\dagger_{x,y} \comp {\biinj_{\faml{Y}}}_y = \coprodM{{\eval_{\prod}}_y \comp
g_x}{\faml{f(x)}_y} \comp {\biinj_{\faml{Y}}}_y = \faml{f(x)}_y$.
\end{enumerate}

To show $g' = g$, suppose $x \in \idx{X}$.
\begin{align*}
\small
g'_x
&=
\lambda_{\prod_{\idx{Y}}}(\reindex{g^\dagger}{(x,-)} \comp \biinj_{\faml{X}_x}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{g^\dagger_{(x,y)} \comp (\biinj_{\faml{X}_x})_y\}_{y \in \idx{Y}}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{\coprodM{{\eval_{\prod}}_y \comp g_x}{\faml{f(x)}_y} \comp (\biinj_{\faml{X}_x})_y\}_{y \in \idx{Y}}) \\
&=
\lambda_{\prod_{\idx{Y}}}(\{{\eval_{\prod}}_y \comp g_x\}_{y \in \idx{Y}}) \\
&=
g_x
\end{align*}
\qed

@@

\begin{center}
\begin{tikzcd}
   \prod_{\idx{Y}}\reindex{\faml{Z}}{\idx{f} \comp (x,-)} \arrow[dr, "\eval_{\prod_{y}}"] \\
   \faml{X}_x \arrow[u, "\lambda_{\prod_{\idx{Y}}}(\reindex{\faml{f}}{{(x,-)}}\,\comp\,
\biinj_{\faml{X}_x})"] \arrow[r, "\faml{f}_{x,y}\,\comp\,\biinj_{\faml{X}_x}"'] & \faml{Z}_{\idx{f}(x,y)}
\end{tikzcd}
\hspace{1cm}
\begin{tikzcd}[column sep=5em]
   \faml{Y}_y \arrow[dr,
   "\faml{f}_{(x,y)}\,\comp\,\biinj_{\faml{Y}_y}"] \\
   \faml{Y}_y \arrow[u, -, double, "\id_{\faml{Y}_y}"] \arrow[r, "\faml{f}_{x,y}\,\comp\,\biinj_{\faml{Y}_y}"'] & \faml{Z}_{\idx{f}(x,y)}
\end{tikzcd}
\end{center}

The diagram on the left below then commutes by Proposition~\ref{prop:biproduct:prod-coprod}, with
$\coprodM{\faml{f}_{x,y}\,\comp\,\biinj_{\faml{X}_x}}{\faml{f}_{x,y}\,\comp\,\biinj_{\faml{Y}_y}} =
\faml{f}_{(x,y)}$ by the uniqueness of the coproduct of morphisms. Then the diagram on the right in $\Fam(C)$
commutes by the definitions of $\lambda_{X,Y,Z}, \eval_{Y,Z}$ and $[Y,Z]$.

\begin{center}
\begin{tikzcd}[column sep=5.5em]
   \prod_{\idx{Y}}\reindex{\faml{Z}}{\idx{f} \comp (x,-)} \biprod \faml{Y}_y \arrow[dr,
   "\coprodM{\eval_{\prod_{y}}}{\faml{f}_{(x,y)}\,\comp\,\biinj_{\faml{Y}_y}}"] \\
   \faml{X}_x \biprod \faml{Y}_y \arrow[u, "(\lambda_{\prod_{\idx{Y}}}(\reindex{\faml{f}}{{(x,-)}}\,\comp\,
\biinj_{\faml{X}_x}){,}\,\id_{\faml{Y}_y})"] \arrow[r,
"\coprodM{\faml{f}_{x,y}\,\comp\,\biinj_{\faml{X}_x}}{\faml{f}_{x,y}\,\comp\,\biinj_{\faml{Y}_y}}"'] &
\faml{Z}_{\idx{f}(x,y)}
\end{tikzcd}
\hspace{1cm}
\begin{tikzcd}
   {{[Y, Z]}} \tensor Y \arrow[dr, "\eval_{Y,Z}"] \\
   X \tensor Y \arrow[u, "\lambda_{X,Y,Z}(f) \tensor \id_Y"] \arrow[r, "f"'] & Z
\end{tikzcd}
\end{center}
\end{enumerate}
\end{proof}
